Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-10 Paper/Draft_harmonization_paper.tex
09.02: Slight edit

\subsection{Harmonising trait databases}

@@

\subsection*{Trait aggregation method selection}

Trait data for freshwater invertebrates is often resolved to a variety of taxonomic levels because species-level identification can be complicated, time-consuming, and expensive (\cite{marshall_taxonomic_2006, resh_which_2008}). Consequently, trait information at family level has been widely used by freshwater ecologists, for example in bioassessment (\cite{beketov_spear_2009}).
However, assigned traits at family level may not reflect the real trait diversity within, for example, a taxon in the large, ecologically diverse family Chironomidae (\cite{serra_synthesising_2016}). Additionally, various trait aggregation methods have been used by researchers to allow comparison among traits at different taxonomic levels. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait is for a particular family. Aggregation of trait information at species or genus level to a point estimate at family level suggests a precision that is not necessarily present, especially not for traits with high variability or if trait information at species level is missing. Indeed, some traits can vary strongly at more precise taxonomic levels than the family level. For example, \citet{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Further, studies focusing on the lability of traits, (i.e. how much traits are constrained by phylogeny), found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and, to a lesser extent, feeding mode are labile, and thus possibly highly variable (\cite{poff_functional_2006, wilkes_traitbased_2020}).

@@

\subsection*{Grouping feature harmonisation}

We explored how using harmonised grouping features and aggregated traits might change the results of trait-environment relationships. We are not aware of other studies that compared trait-environment relationships using harmonised and non harmonised data. The re-analysis of the data on salinisation effects on biological traits from \citet{szocs_effects_2014} yielded only slightly different results compared to the original study, i.e. some of the traits that had responded in the original analysis did not respond in the re-analysis. These non-responsive traits were those closest to the criterion that defined when a trait was associated with either higher or lower salinity, namely feeding mode shredder, respiration gills, and life cycle duration traits. Feeding mode shredder was the only trait that was non-responsive when using harmonised (but not aggregated) as well as when using aggregated data. This result was likely due to the harmonisation procedure, when this trait was amalgamated based on three traits (miner, xylophagus, and shredder). Consequently, the trait affinities in the original data had a higher mean and standard deviation than in the aggregated and harmonised data (Table \ref{tab:SI_resp_traits_summary_stats}), suggesting that the signal in the original data was weakened by the harmonisation. %This might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. -> Kicked out by CM
These findings show that if harmonisation is necessary, harmonised and non-harmonised data, if available, should be compared and possible averaging effects should be considered in further analyses. The fact that the family-level aggregated traits used in the re-analysis showed similar results compared to the original analysis agrees with previous findings that family-level traits can be sensitive enough to detect environmental impacts (\cite{beketov_spear_2009}). 
% Cite another paper that uses family-level trait data
% Although species-level traits are expected to result in more precise results when investigating trait-environment relationships.

@@

Although comprehensive freshwater invertebrate trait databases have been developed for several regions in the past, data synthesis is difficult because of discrepancies in trait definitions. By providing an overview of these definition discrepancies we have set a starting point for the development of standardised trait terminology through which invertebrate trait databases can be linked. A consent on standard terminology and the subsequent development of ontologies are the next steps to facilitate trait-based analysis at large geographical scales. As our analysis showed, some grouping features might need to be re-classified to fit into such a standardised terminology. We could show that trait affinities resulting from fuzzy coding and binary coding can be used together, but a uniform coding of traits is another problem that should be addressed during trait standardisation.

As ecological knowledge increases in the future, and with the aid of new  methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observational and trait data at species-level may become available in the future, reducing the need for trait aggregation. However, if traits need to be aggregated we showed that trait aggregation is often in agreement with expert assignments, especially aggregation approaches using the median. Ultimately, the question which aggregation method is the objectively best measure cannot be answered. Nevertheless, we make the following recommendations for situations when trait aggregation is necessary. 1) The aggregation method will tend to have little effect except for highly variable traits and taxa where a few genera have a high number of species. 2) The coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary coded traits. 3) Weighting trait information based on the taxa present in the database used does not seem to have much influence except for families with few genera that have a high number of species. 4) A measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. 
% OLD: No matter which aggregation approach is chosen, for future studies in which researchers must aggregate traits,....

>>>>

\subsection*{Harmonising trait databases}

@@

\subsection*{Trait aggregation method choice}

Observational data for freshwater invertebrates are often resolved to a variety of taxonomic levels because species-level identification can be complicated, time-consuming, and expensive (\cite{marshall_taxonomic_2006, resh_which_2008}). Consequently, trait information at family level has been widely used by freshwater ecologists, for example in bioassessment (\cite{beketov_spear_2009}).
However, assigned traits at family level may not reflect the real trait diversity within, for example, a taxon in the large, ecologically diverse family Chironomidae (\cite{serra_synthesising_2016}). Additionally, various trait aggregation methods have been used by researchers to allow comparison among traits at different taxonomic levels. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait is for a particular family. Aggregation of trait information at species or genus level to a point estimate at family level suggests a precision that is not necessarily present, especially not for traits with high variability or if trait information at species level is missing. Indeed, some traits can vary strongly at more precise taxonomic levels than the family level. For example, \citet{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Further, studies focusing on the lability of traits, (i.e. how much traits are constrained by phylogeny), found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and, to a lesser extent, feeding mode are labile, and thus possibly highly variable (\cite{poff_functional_2006, wilkes_traitbased_2020}).

@@

\subsection*{Using harmonised and aggregated datasets for ecological research}

By reanalysing trait–environment relationships identified in \citet{szocs_effects_2014} with our harmonised and family-level aggregated datasets, we have illustrated the potential for the use of harmonised and aggregated data in ecological research. As far as we are aware, ours is the first study to compare trait–environment relationships using harmonised and non-harmonised data.
Our re-analysis of the data on salinisation effects on biological traits yielded only slightly different results compared with the original study, which agrees with previous findings that family-level traits can be sensitive enough to detect environmental impacts (\cite{beketov_spear_2009}). However, some of the traits that had responded in the original analysis did not respond in the re-analysis: feeding mode shredder, respiration gills, and life-cycle duration traits. There are several possible reasons for these discrepancies. The non-responsive traits were those closest to the threshold for association with either higher or lower salinity, and so the precision lost through trait harmonisation may have decreased our ability to detect a relatively weak association. Feeding mode shredder was the only trait that was non-responsive when using harmonised (but not aggregated) data as well as when using aggregated data.
This result was likely because of the harmonisation procedure, when this trait was combined based on three traits (miner, xylophagus, and shredder). Consequently, the trait affinities in the original data had a higher mean and standard deviation than in the harmonised and aggregated data (Table \ref{tab:SI_resp_traits_summary_stats}), suggesting that the signal in the original data was weakened by harmonisation. Following this result, we suggest that if harmonisation is necessary, harmonised and non-harmonised data, if available, should be compared, and possible averaging effects should be considered in further analyses. However, our results illustrate that where site-level data are not available, harmonised and aggregated datasets may be useful in identifying at least the strongest associations between traits and environmental conditions.

@@

Although comprehensive freshwater invertebrate trait databases have been developed for several regions in the past, data synthesis presents challenges because of discrepancies in trait definitions, coding of trait affinities, and taxonomic resolution. By providing an overview of definition discrepancies, we have set a starting point for the development of standardised trait terminology through which invertebrate trait databases can be linked. Agreement on standard terminology and the subsequent development of ontologies are the next steps to facilitate trait-based analysis at large geographical scales. As our analysis showed, some grouping features might need to be re-classified to fit into such a standardised terminology. Although we showed that trait affinities from fuzzy and binary coding can be used together, we also recommend that a uniform coding of traits should be addressed during trait standardisation.

As ecological knowledge increases, and with the aid of new  methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observational and species-level trait data at may become available in the future, reducing the need for trait aggregation. However, we showed that trait aggregation is often in agreement with expert assignments, especially through aggregation approaches that use the median.

No aggregation method is objectively best, and except for cases where traits are highly variable and taxonomic hierarchical structure is highly uneven, choice of aggregation method does not appear to be influential on the outcome. Weighting trait information based on the taxa present in the database used does not seem to have much influence except . However, we recommend that the type of coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary-coded traits. We also suggest that a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. Through the use of carefully synthesised data from freshwater invertebrate trait databases, while acknowledging the limitations inherent in harmonised and aggregated data, the ability to investigate a wide range of large-scale ecological questions is enhanced.
